Modelling the Lymphatic Metastatic Progression Pathways of OPSCC from Multi-Institutional Datasets

### 2023-08-29 src/tex/ms.tex
[showyourwork] overleaf sync

% Abstract
\begin{abstract}
    Objective: The \gls{ctv-n}\glsunset{ctv} in \gls{hnscc}\glsunset{scc} is currently based mostly on the prevalence of lymph node metastases in different \glspl{lnl} for a given primary tumor location. In this work, we present a probabilistic model for ipsilateral lymphatic metastatic spread that can quantify the risk for microscopic nodal involvement based on an individual patient's T-stage and clinical involvement of \glspl{lnl} at diagnosis. \\
    Methods: The model is based on the formalism of \glspl{hmm} as conceptually introduced in a previous publication. The \gls{hmm} uses a collection of hidden binary \glspl{rv} -- one for each of the \glspl{lnl} I, II, III, IV, V, and VII -- to model a patient's true state of lymph node involvement corresponding to pathological examinations. Clinical involvement at diagnosis represent the observed binary \glspl{rv} linked to the true state via sensitivity and specificity. A \gls{dag} with is used to compute the transition matrix of the \gls{hmm}, where each edge is associated with a parameter for the probability of tumor spread from the primary tumor and between \gls{lnl}. To learn the parameters, we collected a dataset of 681 \gls{opscc} patients from three institutions. To determine the best graph that balances model complexity and accurate description of \gls{lnl} involvement patterns in the data, we compare graphs based on the model evidence calculated via \gls{ti}. \\
    Results: The graph maximizing the model evidence contains edges for lymphatic spread from \gls{lnl} II to I, III and V, and from \gls{lnl} III to IV and V, and is thus consistent with commonly assumed lymphatic drainage in the neck. Application of the model suggests that the risk of occult disease in level IV is below 5\% if level III is clinically negative, and that the risk of occult disease in level V is below 5\% except for late T-category patients with clinical involvement of levels II, III, and IV. 
    Conclusion: We provide a statistical model of lymph node involvement in \gls{opscc} patients trained on a multi-institutional dataset, which may guide the design of clinical trials on volume-deescalated treatment of \gls{opscc} and eventually contribute to further personalized guidelines on elective nodal treatment.
%    The \gls{ctv-n}\glsunset{ctv} in \gls{hnscc}\glsunset{scc} is currently based mostly on the prevalence of lymphatic metastases in different \glspl{lnl} for a given primary tumor location. In this work, we present an extension to a probabilistic model for lymphatic metastatic spread developed earlier that can quantify the risk for microscopic nodal involvement based on an individual patient diagnosis. The extension is based on the same formalism of \glspl{hmm} as the original model, but in addition to the \glspl{lnl} I, II, III, and IV, it also covers the \glspl{lnl} V and VII of the ipsilateral neck. Moreover, we infer which pathways of lymphatic spread to model from clinical lymphatic progression patterns of 681 patients from three institutions. The extended model may allow for a more personalized \gls{ctv-n} definition based on a patient's individual state of disease. The \gls{hmm} uses a collection of hidden binary \glspl{rv} -- one for each \gls{lnl} -- to model a patient's state of lymphatic involvement. Clinical diagnoses and/or pathological examinations of resected \glspl{lnl} represent the observed binary \glspl{rv} corresponding to the unobservable true state of nodal disease. A \gls{dag} with parametrized edges is used to compute the transition matrix of the \gls{hmm} and consequently the (log-)likelihood of the diagnoses of 681 patients, given a set of spread parameters. Based on the defined likelihood function, \gls{mcmc} sampling is then used to infer a distribution over the model's parameters. T-category is model by assuming that, on average, patients with late T-category tumors are diagnosed at a later time compared to patients with early T-category tumors. Using \gls{ti} and the \gls{bic} we compare models based on different \glspl{dag} and demonstrate the accuracy and precision of the best-performing model. When extended to the contralateral side, this model may form the basis of future guidelines of elective \gls{ctv} definition.
\end{abstract}

@@

%\cite{poortmans_internal_2020,murthy_prostate-only_2021}. 
%Microscopic disease can currently only be observed by a pathological examination of the tissue. Consequently, clinicians that are presented with cancer patients need to routinely assess the risk of microscopic involvement in parts of the body that appear clinically node negative.

This work concerns itself with \gls{opscc}, where approximately 70-80\% of patients present with lymph node metastases at the time of diagnosis.  
%which commonly spreads lymphatically. The only two publicly available datasets on nodal involvement patterns in \gls{opscc}, at the time of writing, show that 80\% of patients where either cN+ or -- if available -- pN+ \cite{ludwig_dataset_2022}. Since the exact number and location of lymph nodes varies from patient to patient, the nodes are grouped into anatomically defined \glspl{lnl}. These levels are then often irradiated or resected prophylactically due to the risk of harboring occult metastases despite negative findings from imaging modalities. 
In clinical practice, \gls{ctv-n} definition in radiotherapy is based on guidelines \cite{gregoire_ct-based_2003,gregoire_delineation_2014,gregoire_delineation_2018,eisbruch_intensity-modulated_2002,biau_selection_2019,chao_determination_2002,vorwerk_guidelines_2011,ferlito_elective_2009} that are mostly derived from the observed prevalence of involvement in an \gls{lnl} for a given tumor location. These guidelines currently suggest extensive irradiation of both sides of the neck for most patients. In the ipsilateral neck, the \gls{ctv-n} includes \glspl{lnl} II, III and IV for all patients, and levels I and V for the majority of patients. These guidelines, however, do not account for the personal risk of the patients that may depend greatly on their state of tumor progression at diagnosis. E.g., a patient with macroscopic metastases detected via \gls{pet} in both \glspl{lnl} II and III may have a substantial risk for occult disease in \gls{lnl} IV. Instead, patients who present with a clinically N0 neck or a single metastasis in \gls{lnl} II may have a much smaller risk for occult disease in \gls{lnl} IV.

We previously developed a model of lymphatic metastatic progression for estimating the risk of microscopic disease, given a patient's personal diagnosis. The initial model was based on the methodology of Bayesian networks \cite{pouymayou_bayesian_2019}. It was subsequently extended and formulated as a \glspl{hmm} to include T-category in an intuitive manner \cite{ludwig_hidden_2021}. However, these models were introduced based on only a small dataset of approximately 100 early T-category patients \cite{sanguineti_defining_2009} available at that time. The limited data did not allow us to quantify the probability of metastases in the rarely involved \gls{lnl} I, V, and VII, nor did the data allow us to verify that the \gls{hmm} is adequate to describe the dependence on lymph node involvement on T-category.
%Since this publication only includes details about the involvement of the ipsilateral \glspl{lnl} I, II, III, and IV, the models were limited to these nodal levels as well.
%With the publication of larger and more detailed datasets \cite{ludwig_dataset_2022}, we are thus able to
In this paper, we extend the previous work \cite{ludwig_hidden_2021} by making the following contributions:
\begin{enumerate}
    \item We provide a \glspl{hmm} of ipsilateral lymph node involvement including all relevant \glspl{lnl}, namely the levels I, II, III, IV, V, and VII. To determine the underlying \gls{dag} we compare different graphs by calculating the model evidence through \gls{ti}.
    \item We collect a multi-centric dataset consisting of \variable{output/num_patients.txt} patients from three institutions, allowing us to train the model based on a sizable dataset.
    \item We use the trained model to provide personalized risk estimations for occult metastases for typical clinical states of tumor progression at diagnosis, illustrating its potential for guiding volume-deescalated treatment strategies in the future. 
\end{enumerate}

@@

The terms of the form $P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)$ implicitly depend on how we parametrize e.g. the arcs of \cref{fig:graph_with_obs}. For example, if we look at the probability of spread to the \gls{lnl} $X_3$ depending on the state of that level's parent -- which, in this case, is $\operatorname{pa}(3) = 2$ -- we can write the different combinations into a conditional probability table as below.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{figures/graph_with_obs.png}
    \caption{\Gls{dag} representing a possible abstraction of the lymphatic network comprising the tumor (red shaded circle) and \glspl{lnl} II through IV as hidden binary \glspl{rv} (blue outlined circles). Attached to each of these is the corresponding observed \gls{rv} (orange shaded squares). Lymphatic flow is depicted in the form of parametrized arrows (red and blue) that represent the probability of spread along the respective arc per time-step. Sensitivity and specificity (orange arrows) connect the hidden \glspl{rv} to the diagnosis.}
    \label{fig:graph_with_obs}
\end{figure}

@@

\ju{Also say that the above equation gives the probability of a hidden state. To get the probability of involvement in a given level one has to marginalize over all corresponding states}

@@

%I wouldn't even describe these minor deviations

%There are only a few involvement patterns for which the model's prediction does not match the prevalence of the data with perfect accuracy. But in these instances, e.g. the case where \gls{lnl} III is metastatic while \gls{lnl} IV is healthy in an early T-category patient (\cref{fig:bg_prevalences}, third row, green line and histogram), the model slightly overestimates the prevalence. An overestimation by a few percent however, does not pose a risk to the model's clinical applicability, since in the context of volume-deescalation, this can be considered cautious.

@@

\ju{perhaps one could use the same color in each risk figure for a given clinical involvement, e.g. always green for N0}
\rl{Done}

\begin{figure}
    \script{wg_III_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_III_risks.png}
        \caption{TODO: Write caption.}
        \label{fig:wg_III_risks}
    \end{centering}
\end{figure}

@@

\begin{figure}
    \script{wg_any_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_any_risks.png}
        \caption{TODO: Shown are the historgams over the predicted risk for involvement in any of the \glspl{lnl} I, IV, V, or VII. The risk is plotted given a clinical N0 diagnosis (green), macroscopically detected metastases in \gls{lnl} II (blue), and lastly given visible involvement in both \gls{lnl} II and III (orange). The top row shows these risks for early T-category diagnoses and the bottom row for late T-category.}
        \label{fig:wg_any_risks}
    \end{centering}
\end{figure}

@@

\subsection{Implications for elective nodal treatment} 
Risk predictions obtained by the model may be used to design clinical trials on volume-deescalated treatment of \gls{opscc}. In the context of radiotherapy, this corresponds to excluding \glspl{lnl} from the CTV-N, which are irradiated according to the current guidelines. The list below should be seen as a summary of \cref{subsec:results:risk_prediction} and the limitations discussed in \cref{subsec:disc:limitation} should be taken into account in its interpretation. Assuming that one accepts approximately a 5\% risk of occult metastases per \gls{lnl}, the statistical model presented in this paper would suggest to:
\begin{itemize}
    \item[$\bullet$] Irradiate level II for all patients. \ju{maybe we should actually show the risk in level II for N0 patients, at least state the risk in the text, are even have the risk histogram for completeness.}
    \item[$\bullet$] Irradiate level III for most patients. Only for clinically N0, early T-category patients, not irradiating level III could be considered. 
    \item[$\bullet$] Exclude level IV from the CTV-N for patients with clinically negative level III. For advanced T-category patients with involvement of level III, level IV should be irradiated. %For early T-category patients with involvement of level III, not irradiating level IV could be considered.
    \item[$\bullet$] Exclude level V from the CTV-N for most patients. Only for patients with extensive involvement of levels II, III, and IV, irradiation of level V can be considered.
    \item[$\bullet$] Exclude level I from the CTV-N for early T-category patients with limited metastatic disease. For advanced T-category patients with extensive nodal involvement of levels II, III, and IV, level I should be irradiated (see also the limitations discussed below).
    \item[$\bullet$] \ju{Level VII}
\end{itemize}

@@

>>>>

% Abstract
\begin{abstract}
    Objective: The \gls{ctv-n}\glsunset{ctv} in \gls{hnscc}\glsunset{scc} is currently based mostly on the prevalence of lymph node metastases in different \glspl{lnl} for a given primary tumor location. In this work, we present a probabilistic model for ipsilateral lymphatic metastatic spread that can quantify the risk for microscopic nodal involvement based on an individual patient's T-stage and clinical involvement of \glspl{lnl} at diagnosis. \\
    Methods: The model is based on the formalism of \glspl{hmm} as conceptually introduced in a previous publication. The \gls{hmm} uses a collection of hidden binary \glspl{rv} -- one for each of the \glspl{lnl} I, II, III, IV, V, and VII -- to model a patient's true state of lymph node involvement corresponding to pathological examinations. Clinical involvement at diagnosis represent the observed binary \glspl{rv} linked to the true state via sensitivity and specificity. A \gls{dag} with is used to compute the transition matrix of the \gls{hmm}, where each edge is associated with a parameter for the probability of tumor spread from the primary tumor and between \gls{lnl}. To learn the parameters, we collected a dataset of 681 \gls{opscc} patients from three institutions. To determine the best graph that balances model complexity and accurate description of \gls{lnl} involvement patterns in the data, we compare graphs based on the model evidence calculated via \gls{ti}. \\
    Results: The graph maximizing the model evidence contains edges for lymphatic spread from \gls{lnl} II to I, III and V, and from \gls{lnl} III to IV and V, and is thus consistent with commonly assumed lymphatic drainage in the neck. Application of the model suggests that the risk of occult disease in level IV is below 5\% if level III is clinically negative, and that the risk of occult disease in level V is below 5\% except for late T-category patients with clinical involvement of levels II, III, and IV. 
    Conclusion: We provide a statistical model of lymph node involvement in \gls{opscc} patients trained on a multi-institutional dataset, which may guide the design of clinical trials on volume-deescalated treatment of \gls{opscc} and eventually contribute to further personalized guidelines on elective nodal treatment.
\end{abstract}

@@

This work concerns itself with \gls{opscc}, where approximately 70-80\% of patients present with lymph node metastases at the time of diagnosis.   In clinical practice, \gls{ctv-n} definition in radiotherapy is based on guidelines \cite{gregoire_ct-based_2003,gregoire_delineation_2014,gregoire_delineation_2018,eisbruch_intensity-modulated_2002,biau_selection_2019,chao_determination_2002,vorwerk_guidelines_2011,ferlito_elective_2009} that are mostly derived from the observed prevalence of involvement in an \gls{lnl} for a given tumor location. These guidelines currently suggest extensive irradiation of both sides of the neck for most patients. In the ipsilateral neck, the \gls{ctv-n} includes \glspl{lnl} II, III and IV for all patients, and levels I and V for the majority of patients. These guidelines, however, do not account for the personal risk of the patients that may depend greatly on their state of tumor progression at diagnosis. E.g., a patient with macroscopic metastases detected via \gls{pet} in both \glspl{lnl} II and III may have a substantial risk for occult disease in \gls{lnl} IV. Instead, patients who present with a clinically N0 neck or a single metastasis in \gls{lnl} II may have a much smaller risk for occult disease in \gls{lnl} IV.

We previously developed a model of lymphatic metastatic progression for estimating the risk of microscopic disease, given a patient's personal diagnosis. The initial model was based on the methodology of Bayesian networks \cite{pouymayou_bayesian_2019}. It was subsequently extended and formulated as a \glspl{hmm} to include T-category in an intuitive manner \cite{ludwig_hidden_2021}. However, these models were introduced based on only a small dataset of approximately 100 early T-category patients \cite{sanguineti_defining_2009} available at that time. The limited data did not allow us to quantify the probability of metastases in the rarely involved \gls{lnl} I, V, and VII, nor did the data allow us to verify that the \gls{hmm} is adequate to describe the dependence on lymph node involvement on T-category. In this paper, we extend the previous work \cite{ludwig_hidden_2021} by making the following contributions:
\begin{enumerate}
    \item We provide a \glspl{hmm} of ipsilateral lymph node involvement including all relevant \glspl{lnl}, namely the levels I, II, III, IV, V, and VII. To determine the underlying \gls{dag} we compare different graphs by calculating the model evidence through \gls{ti}.
    \item We collect a multi-centric dataset consisting of \variable{output/num_patients.txt} patients from three institutions, allowing us to train the model based on a sizable dataset.
    \item We use the trained model to provide personalized risk estimations for occult metastases for typical clinical states of tumor progression at diagnosis, illustrating its potential for guiding volume-deescalated treatment strategies in the future. 
\end{enumerate}

@@

The terms of the form $P \left( \xi_{iv} \mid \left\{ \xi_{jr} \right\}_{r \in \operatorname{pa}(v)} \right)$ implicitly depend on how we parameterize e.g. the arcs of \cref{fig:graph_with_obs}. For example, if we look at the probability of spread to the \gls{lnl} $X_3$ depending on the state of that level's parent -- which, in this case, is $\operatorname{pa}(3) = 2$ -- we can write the different combinations into a conditional probability table as below.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{figures/graph_with_obs.png}
    \caption{\Gls{dag} representing a possible abstraction of the lymphatic network comprising the tumor (red shaded circle) and \glspl{lnl} II through IV as hidden binary \glspl{rv} (blue outlined circles). Attached to each of these is the corresponding observed \gls{rv} (orange shaded squares). Lymphatic flow is depicted in the form of parameterized arrows (red and blue) that represent the probability of spread along the respective arc per time-step. Sensitivity and specificity (orange arrows) connect the hidden \glspl{rv} to the diagnosis.}
    \label{fig:graph_with_obs}
\end{figure}

@@

From this probability for any possible hidden state, we can also compute the probability of, for example, involvement in \gls{lnl} IV. To that end, we marginalize over all states $\boldsymbol{\xi}_k$ where $\xi_{k4} = 1$, meaning those states in which \gls{lnl} IV harbors metastases. Formally, we can define a marginalization vector $\mathbf{m}$ that is one for every hidden state we want to include in the marginalization and zero elsewhere. In the example of the marginalized probability for \gls{lnl} IV involvement, the components would look like this:
%
\begin{equation}
    m_{4k} = \operatorname{id}(\xi_{k4} = 1)
\end{equation}
%
Subsequently, we can compute the marginalization as a dot product:
%
\begin{equation}
    \begin{aligned}
        P\left(\text{IV} = 1 \mid \mathbf{Z}=\boldsymbol{\zeta}_\ell \right) &= \sum_{k:\xi_{k4} = 1} P\left( \mathbf{X}=\boldsymbol{\xi}_k \mid \mathbf{Z}=\boldsymbol{\zeta}_\ell \right) \\
        &= \mathbf{m}_4 \cdot P\left( \mathbf{X}=\boldsymbol{\xi}_k \mid \mathbf{Z}=\boldsymbol{\zeta}_\ell \right)
    \end{aligned}
\end{equation}
%
% \ju{Also say that the above equation gives the probability of a hidden state. To get the probability of involvement in a given level one has to marginalize over all corresponding states}

@@

@@

% \ju{perhaps one could use the same color in each risk figure for a given clinical involvement, e.g. always green for N0}

\begin{figure}
    \script{wg_III_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_III_risks.png}
        \caption{Histograms over the risk for microscopic involvement in \gls{lnl} III, given that a patient presents as clinically N0 (green), or given that \gls{lnl} II shows clinical involvement (blue). The top panel shows these risks for early T-category, the bottom panel for late.}
        \label{fig:wg_III_risks}
    \end{centering}
\end{figure}

@@

\begin{figure}
    \script{wg_any_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_any_risks.png}
        \caption{Shown are the histograms over the predicted risk for involvement in any of the \glspl{lnl} I, IV, V, or VII. The risk is plotted given a clinical N0 diagnosis (green), macroscopically detected metastases in \gls{lnl} II (blue), and lastly given visible involvement in both \gls{lnl} II and III (orange). The top row shows these risks for early T-category diagnoses and the bottom row for late T-category.}
        \label{fig:wg_any_risks}
    \end{centering}
\end{figure}

@@

\subsection{Implications for elective nodal treatment} 
Risk predictions obtained by the model may be used to design clinical trials on volume-deescalated treatment of \gls{opscc}. In the context of radiotherapy, this corresponds to excluding \glspl{lnl} from the CTV-N, which are irradiated according to the current guidelines. The list below should be seen as a summary of \cref{subsec:results:risk_prediction} and the limitations discussed in \cref{subsec:disc:limitation} should be taken into account in its interpretation. Assuming that one accepts approximately a 5\% risk of occult metastases per \gls{lnl}, the statistical model presented in this paper would suggest to:
\begin{itemize}
    \item[$\bullet$] Irradiate level II for all patients.%\ju{maybe we should actually show the risk in level II for N0 patients, at least state the risk in the text, are even have the risk histogram for completeness.}
    \item[$\bullet$] Irradiate level III for most patients. Only for clinically N0, early T-category patients, not irradiating level III could be considered. 
    \item[$\bullet$] Exclude level IV from the CTV-N for patients with clinically negative level III. For advanced T-category patients with involvement of level III, level IV should be irradiated.
    \item[$\bullet$] Exclude level V from the CTV-N for most patients. Only for patients with extensive involvement of levels II, III, and IV, irradiation of level V can be considered.
    \item[$\bullet$] Exclude level I from the CTV-N for early T-category patients with limited metastatic disease. For advanced T-category patients with extensive nodal involvement of levels II, III, and IV, level I should be irradiated (see also the limitations discussed below).
    \item[$\bullet$] Exclude level VII in all patients, unless it is clinically involved.
\end{itemize}

@@

\section{Acknowledgments}

This work was supported by the Swiss Cancer Research Foundation under grant \href{https://www.krebsforschung.ch/unterstuetzen-sie-uns/stiftungen/-dl-/fileadmin/downloads/unterstuetzen-sie-uns/projekte-der-stiftung-krebsforschung-schweiz-2023.pdf}{KFS-5645-08-2022} and by the University Zürich under the Clinical Research Priority Program \href{https://www.crpp-ai-oncology.uzh.ch/en/Projects/Project-5.html}{Artificial Intelligence in Oncological Imaging}.
